Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-02-26 papers/wdmerger_I/paper.tex
Change default initial model resolution to 25 km and make it a probin parameter. Add discussion of this choice to the paper.

We desire to specify the mass of the white dwarf, as well as its
temperature and composition. To start the integration off, we
therefore need to guess at a central density.  We then do a secant
iteration over the entire integration procedure to find the central
density needed to yield the desired total mass.  We note that the grid
spacing $\Delta x$ is chosen such that it is equal to the zone size on
the finest level of our problem, which is known in advance since we
specify the domain size and the maximum number of possible levels of
refinement.

We map the 1D model onto the 3D Cartesian grid by taking density,
temperature, and composition as the independent variables,
interpolating these to the cell centers, and then calling the equation
of state to initialize the remaining terms.  The interpolation process
divides each zone into $n_{\text{sub}}$ sub-zones of equal volume for
the purpose of sampling the 1D model, and the sub-zones are added
together to obtain the full zone's characteristics. This
sub-grid-scale interpolation is useful near the edge of the star,
where the density falls off rapidly with radius.

@@

The initial velocity is taken to be zero in if we are in the reference
frame that rotates with the WDs, and if we are in the inertial frame
the velocity is set equal to the rigid rotation rate corresponding to
the specified period $T$.

We do not attempt to enforce equilibrium with an additional relaxation
step. This will be an important part of future work in this series, as
numerous groups working on binary evolution
\citep{swc:2000,motl:2002,rosswog:2004,dan:2011,pakmor:2012:gadget}
have commented on the importance of equilibrium initial conditions in
determining the evolution of the system. In particular, we plan to
compare the self-consistent field method (which finds an equilibrium
solution \textit{a priori}) and the method of \cite{rosswog:2004},
which starts with our method and then attempts to guide the evolution
toward the equilibrium state using suitably defined damping
terms. However, these issues are not considered for the present study,
as we are not yet attempting to study the merger process.

As a consequence of starting in a non-equilibrium setting, there are 
large density and pressure gradients near the white dwarf surfaces
that result in large amounts of mass flowing into the ambient
medium. This can result in spurious non-physical consequences such as 
the total density or energy going negative in a zone. To compensate 
for this, we start the simulation with a timestep that is a few orders 
of magnitude smaller than that required by the CFL criterion, and allow
the timestep to increase by 1\% each timestep so that the timestep reaches 
its maximum allowed by the velocities on the grid over a span of approximately 
1000 timesteps. This allows the ambient medium just outside the white dwarf
to come closer to equilibrium with the surface without having 
discontinuous jumps in the density or energy. For all simulations, 
the maximum timestep is set to be equal to one-half of the CFL limit.

@@

>>>>

We desire to specify the mass of the white dwarf, as well as its
temperature and composition. To start the integration off, we
therefore need to guess at a central density.  We then do a secant
iteration over the entire integration procedure to find the central
density needed to yield the desired total mass.  The grid spacing is
$\Delta x = 25\ \text{km}$. We chose this value because no simulation
we perform is likely to exceed this grid resolution inside the stars 
themselves; for our normal domain size (see below), this corresponds to 
three jumps in refinement by a factor of four. We find that for low 
resolution runs, this is a better choice than selecting the 1D grid 
spacing to be equal to the 3D grid spacing.

We map the 1D model onto the 3D Cartesian grid by taking density,
temperature, and composition as the independent variables,
interpolating these to the cell centers, and then calling the equation
of state to initialize the remaining terms. It is possible to interpolate
instead by using pressure instead of temperature, as pressure is more 
closely related to hydrostatic balance, but the EOS we use is so 
insensitive to temperature that this mapping can result in large 
deviations from the isothermal assumption we started with.  The 
interpolation process divides each zone into $n_{\text{sub}}$ 
sub-zones of equal volume for
the purpose of sampling the 1D model, and the sub-zones are added
together to obtain the full zone's characteristics. This
sub-grid-scale interpolation is useful especially near the edge of the star,
where the density falls off rapidly with radius. Typically we take 
$n_{\text{sub}} = 4$.

@@

The initial velocity is taken to be zero in if we are in the reference
frame that rotates with the WDs, and if we are in the inertial frame
the velocity in every zone is set equal to the rigid rotation rate 
corresponding to the distance of that zone from the rotation axis, given
the specified period $T$. Thus the inertial frame and rotating frame 
simulations are starting off with the same initial conditions: two white 
dwarfs locked in synchronous rotation. This is the simplest assumption to 
make, but in the future we may explore relaxing this requirement.

We do not attempt to enforce equilibrium with an additional relaxation
step. This will be an important part of future work in this series, as
numerous groups working on binary evolution
\citep{swc:2000,motl:2002,rosswog:2004,dan:2011,pakmor:2012:gadget}
have commented on the importance of equilibrium initial conditions in
determining the evolution of the system. As a consequence of starting 
in a non-equilibrium setting, there are 
large density and pressure gradients near the white dwarf surfaces
that result in large amounts of mass flowing into the ambient
medium. This can result in spurious non-physical consequences such as 
the total density or energy going negative in a zone. To compensate 
for this, we start the simulation with a timestep that is a few orders 
of magnitude smaller than that required by the CFL criterion, and allow
the timestep to increase by 1\% each timestep so that the timestep reaches 
its maximum allowed by the velocities on the grid over a span of approximately 
1000 timesteps. This allows the ambient medium just outside the white dwarf
to come closer to equilibrium with the surface without having 
discontinuous jumps in the density or energy. For all simulations, 
the maximum timestep is set to be equal to one-half of the CFL limit.

@@

The computational domain has a total size of $1.024 \times 10^{10}\ \text{cm}$ in each 
spatial dimension, and is centered at the origin. Our coarse grid has $256^3$ zones,
corresponding to a spatial resolution of 400 km. For the present study, we choose a simple 
refinement strategy: all zones within the Roche radius of each star are tagged for maximal 
refinement. We also ensure that the outer part of the domain is never tagged for refinement.
In future work we will add criteria that tag for refinement the gas between the stars, which is
expected to feature nuclear burning.
